---
title: "My example of ggtree to visualize phylogeny"
output: html_notebook
author: Pengyao Jiang
date: 06.25.2020
---


### For more information about the ggtree R package, check ggtree online [documentation](https://guangchuangyu.github.io/ggtree-book/chapter-ggtree.html)
### You can find data and the source of R notebook from [here](https://github.com/pyjiang/ggtree_example). The distance matrix is from [Peter et al, Nature 2018](https://www.nature.com/articles/s41586-018-0030-5) and the annotation is adapted from a subset of the orignal data.  


#### load pacakges 
```{r}
library(ape)
library(phangorn)
library(ggplot2)
library(ggtree)
library(colorspace)
library(tibble)
library(tidytree)
library(treeio)
```

#### read in the distance matrix of 1011 strains
```{r}
## use phylogenetic tree built from SNPs 
y1002_dist <- read.table("1011DistanceMatrixBasedOnSNPs.tab",header=TRUE,row.names=1)

head(y1002_dist)
```

#### read the annotation file with a subset of 618 strains 
```{r}
annot_new <- read.table("strain_annot_new.csv", header =TRUE, sep=",",
                    stringsAsFactors = FALSE)


head(annot_new)
```


#### get the subset of the distance matrix with 618 strains, then using NJ to build a tree

```{r}

all_strains <- names(y1002_dist)
subset_strains <- annot_new[,"ID"]


subset <- all_strains[all_strains %in% subset_strains]

y1002_dist_sub <- y1002_dist[subset,subset]


### generate a NJ tree by this subset of strains
### add pseudo root with CEG

y1002_sub_nj <- NJ(as.matrix(y1002_dist_sub))

y1002_sub_nj_new <- root(y1002_sub_nj, "CEG")
```

#### if visualizing the raw tree  
```{r}
ggtree(y1002_sub_nj_new)
```

#### Or in the circular form. 


```{r}
ggtree(y1002_sub_nj_new, layout="circular")

```


#### The raw trees do not look pretty. Try to see some features from ggtree with a subset of the orinal tree. 
#### Get a subtree.
```{r}
## only get subset of beer strains
strains <- c("BRM", "SACE_YAG", "AEQ", "AAR", "SACE_YAB", "AQH", "AFP")


## find the node for common ancester of this clade 
node_show <- MRCA(y1002_sub_nj_new, strains)

## subset the original tree 
# https://tbradley1013.github.io/2018/06/19/subsetting-phylogenetic-trees/
# https://rdrr.io/github/GuangchuangYu/treeio/man/tree_subset.html
subtree <- tree_subset(y1002_sub_nj_new, node = node_show, levels_back =0 )

subtree
```

```{r}

## phylo object can be easily converted to tibble (easy for annotation)

x <-as_tibble(subtree)

## and can be converted back to phylo, using as_phylo()
## as.phylo(x)

## also can be converted to tree data, which is a phylo object but with annotation data (in theory)
## https://guangchuangyu.github.io/2017/01/convert-graphic-object-to-tree-object-using-treeio/
## as.treedata(x)
## but I was having problems using this function where the annotation seems to be lost in the conversion.


x
```


#### Annotate the tree: assign strains to 2 groups 

```{r}

## define two clades, one has enrichment of C>A, the other don't
groups <- list(c1= c("BRM", "SACE_YAG", "AEQ", "AAR"),
               c2 = c("SACE_YAB", "AQH", "AFP","AFA", "AFB"))

subtree2 <- groupOTU(subtree, groups)

p <- ggtree(subtree2)+geom_tiplab( size=4)
p
```

#### scale the tree for output
```{r}

p <- p+ xlim(0,0.7)

p
```



#### If want a circular tree
```{r}
ggtree(subtree2, layout= "circular")+geom_tiplab( size=4)

```


#### add a circle to strains with two different colors indicating the group information 
```{r}
p + geom_tippoint(aes(color=group), size=3) +scale_color_manual(values=c("#1c7c54","#999999")) 
```

#### or color the strain names
```{r}

ggtree(subtree2)+geom_tiplab( aes(color=group), size=4) + xlim(0,0.7)

```


#### Collapse a clade.  (for AFP, AFA, AFB)


```{r}
## find common ancester of AFP, AFA, AFB
clade1 <- MRCA(subtree2, c("AFP", "AFB"))

clade1
```

#### collapse 
```{r}
collapse(p, clade1, 'min', fill="#999999")  
```

#### scale the collapsed clade
```{r}

## need to scale first, and then collapse 

p3 <- scaleClade(p, clade1, 0.3)
collapse(p3, clade1, 'min', fill="#999999")  
```

#### Or highlight a clade 

```{r}
p + geom_hilight(node=clade1, fill="#2F30EE", alpha=0.45)
```

#### If want to show a bunch of strains, which do not belong to a monophyletic group, use the function `geom_strip`

```{r}
p + geom_strip("SACE_YAB","AFB",barsize=1,color="red", label = "group X") 
```





#### A real example of a coloring a whole phylogenetic tree 

```{r}
## this example used the original phylogentic tree (y1002_sub_nj_new) 
## but only use 3 columns from the annotation file to annotate: CAgroup, Clades, group


## only keep Clades and group
subtb_annot2 <- annot_new[,c("ID", "CAgroup", "Clades", "group")]


names(subtb_annot2) <- c("label", "CAgroup", "Clades", "group")

subtb_annot2new <- subtb_annot2[,c("label", "CAgroup")]

subtb_annot2new[,"CAgroup"] <- factor(subtb_annot2new[,"CAgroup"])


## then add group information by clade 

## exclude strains which are not assignned to a clade
subtb_annot3 <- subtb_annot2[!is.na(subtb_annot2[,"Clades"]),]

## trim spaces 
subtb_annot3[,"Clades"] <- trimws(subtb_annot3[,"Clades"])

names(subtb_annot3) <- c("label", "CAgroup", "Clades", "group")


## https://bioconductor.riken.jp/packages/3.4/bioc/vignettes/ggtree/inst/doc/treeManipulation.html
## generate a list of tip nodes, to feed into ggtree, to color sub-branches
clades <- levels(factor(subtb_annot3[,"Clades"]))
nclades <- length(clades)
clades_for_ggtree <- list()
for(i in 1:nclades)
{
  clade_name <- clades[i]
  subset <- subtb_annot3 [subtb_annot3[,"Clades"] == clade_name,"label"]
  clades_for_ggtree[[clade_name]] <- subset
}



tree <- groupOTU(as_tibble(y1002_sub_nj_new), clades_for_ggtree)

## add annotation of CA group
tree2 <- as.treedata(full_join(tree, subtb_annot2new, by="label"))



## find the node for each clade 
## 1. wine 
n1 <- MRCA(tree2, c("ATM", "AHD"))
## 3. brazilian ethanol
n2 <- MRCA(tree2, c("BVC", "CNL"))
## 4. medteranina oak
n3 <- MRCA(tree2, c("BFP", "CCL"))
## 5. french dairy 
n4 <- MRCA(tree2, c("ARS", "BGE"))
## 6. african beer 
n5 <- MRCA(tree2, c("ANM", "AFL"))
## 9.mexican agave
n6 <- MRCA(tree2, c("CPL", "CPN"))
## 10. french guaina human 
n7 <- MRCA(tree2, c("BNB", "BDQ"))
## 11. ale beer 
n8 <- MRCA(tree2, c("CGC", "CFC"))
## 12. west african coca
n9 <- MRCA(tree2, c("CQN", "CQD"))
## 13. african palm wine 
n10 <- MRCA(tree2, c("AKH", "BAA"))

## 24. asian islands 
n11 <- MRCA(tree2, c("CCV", "ARH"))
## 25. sake
n12 <- MRCA(tree2, c("ARP", "ANR"))
## 26. asian ferm 
n13 <- MRCA(tree2, c("AKL", "AHQ"))

## Taiwanese
treetb <- as_tibble(tree2)
n14_pre <- treetb[treetb[,"label"] == "CEG","node"]
n14 <- n14_pre[!is.na(n14_pre)]

## 7. mosaic beer 
n15 <- MRCA(tree2, c("SACE_YAG", "CBN"))
## 8. mixed origin 
n16 <- MRCA(tree2, c("CHG", "AFF"))



## zoom in to scale the clade that has specific strains 
z1 <- MRCA(tree2, c("SACE_YAG", "BRM"))
z2 <- MRCA(tree2, c("CHG", "AFF"))
z3 <- MRCA(tree2, c("ADT", "ACN"))


## originally used group field to color, therefore have so many different colors 
## now only use CAgroup field to color tip labels. (all the clade names and their colors are not used.)
color_vec <- c( "1. Wine/European"= "#5B1A18", 
                "1. Wine/European (subclade 1)"= "#5B1A18",
                "1. Wine/European (subclade 2)"= "#5B1A18",
                "1. Wine/European (subclade 4)" = "#5B1A18",
                "2. Alpechin" = "#f76a8c",
                "3. Brazilian bioethanol"="#f76a8c",
                "4. Mediterranean oak" = "#377eb8",
                "5. French dairy"= "#fcae91", 
                "6. African beer"=  "#377eb8", 
                "7. Mosaic beer"="#6ADD77", 
                "8. Mixed origin"= "#4C00A3",
                "9. Mexican agave" = "#6ADD77",
                "10. French Guiana human" = "#fcae91",
                "11. Ale beer" = "#FAD510",
                "12. West African cocoa" = "#FAD510",  
                "13. African palm wine"= "#9c5518",
                "14. CHNIII" = "#CC6677",
                "15. CHNII" = "#CC6677",
                "16. CHNI" = "#CC6677",
                "17. Taiwanese" =  "#882255",
                "18. Far East Asia" = "#CC6677",
                "19. Malaysian" = "#CC6677",
                "20. CHN V" = "#CC6677",
                "21. Ecuadorean" = "#CC6677",
                "22. Far East Russian" = "#CC6677",
                "23. North American oak"="#3CAEA3",
                "24. Asian islands"= "#fbcffc",
                "25. Sake" = "#be79df", 
                "26. Asian fermentation"= "#2b580c",
                "M1. Mosaic region 1" = "#be79df",
                "M2. Mosaic region 2" = "#f76a8c",
                "M3. Mosaic region 3" = "#87bbd6",
                "0" = "#999999",
                "1" = "#C70405"
)



p <- ggtree(tree2, layout= "circular", size=0.3)

p <- rotate_tree(p, 180)

## only strains that has CAgroup (non NA value, should have a value of 1) will be shown of tip labels in red
## CAgroup needs to be factor 
p <- p + geom_tiplab(aes(color= CAgroup),  size=2, hjust = 1)  +
  scale_color_manual(values=color_vec)



p <- scaleClade(p, n1, 0.3)
p <- scaleClade(p, n2, 0.5)
p <- scaleClade(p, n3, 0.65)
p <- scaleClade(p, n4, 0.5)
p <- scaleClade(p, n5, 0.5)
p <- scaleClade(p, n6, 0.5)
p <- scaleClade(p, n7, 0.5)
p <- scaleClade(p, n8, 0.5)
p <- scaleClade(p, n9, 0.5)
p <- scaleClade(p, n10, 0.5)
p <- scaleClade(p, n11, 0.5)
p <- scaleClade(p, n12, 0.5)
p <- scaleClade(p, n13, 0.5)

p <- scaleClade(p, z1, 5.2)
p <- scaleClade(p, z2, 5.2)
p <- scaleClade(p, z3, 5.2)

```


```{r}


p2 <- collapse(p, n1, 'min', fill="#999999")  

p2 <- collapse(p2, n2, 'min', fill="#999999")  

p2 <- collapse(p2, n3, 'min', fill="#999999")  

p2 <-  collapse(p2, n4, 'min', fill="#999999") 

p2 <- collapse(p2, n5, 'min', fill="#999999")  

p2 <- collapse(p2, n6, 'min', fill="#999999")  

p2 <- collapse(p2, n7, 'min', fill="#999999")  

p2 <-  collapse(p2, n8, 'min', fill="#999999") 

p2 <- collapse(p2, n9, 'min', fill="#999999")  

p2 <- collapse(p2, n10, 'min', fill="#999999")  

p2 <-  collapse(p2, n11, 'min', fill="#999999") 

p2 <- collapse(p2, n12, 'min', fill="#999999")  

p2 <-  collapse(p2, n13, 'min', fill="#999999") 




p2 <- p2 + geom_cladelabel(n1, "Wine/European", fontsize=3, offset=0, hjust=1.2) +
  geom_cladelabel(n2, "Brazilian bioethanol", angle=90, fontsize=3, offset=0.7, hjust=0.3) +
  geom_cladelabel(n3, "Mediterranean oak",  angle=90, fontsize=3, offset=0.5, hjust=0) +
  geom_cladelabel(n4, "French dairy", angle=90, fontsize=3, offset=0.8, hjust=0.5) +
  geom_cladelabel(n5, "African beer", angle=90, fontsize=3, offset=0.6, hjust=0.3) +
  geom_cladelabel(n15, "Mosaic beer",angle=10, fontsize=3, offset=0.5, hjust=1, barsize=0.5) + 
  geom_cladelabel(n16, "Mixed origin",angle=30, fontsize=3, offset=0.5, hjust=1, barsize=0.5) + 
  geom_cladelabel(n6, "Mexican agave",angle=300, fontsize=3, offset=0.8, hjust=0.5) + 
  geom_cladelabel(n7, "French Guiana human",angle=300, fontsize=3, offset=0.8, hjust=0.5) + 
  geom_cladelabel(n8, "Ale beer",angle=300, fontsize=3, offset=0.8, hjust=1) + 
  geom_cladelabel(n9, "West African cocoa",angle=270, fontsize=3, offset=0.3, hjust=0) + 
  geom_cladelabel(n10, "African palm wine",angle=10, fontsize=3, offset=0.9, hjust=0.5) + 
  geom_cladelabel(n11, "Asian islands",angle=10, fontsize=3, offset=0.9, hjust=0.5) + 
  geom_cladelabel(n12, "Sake",angle=40, fontsize=3, offset=0.2, hjust=1) + 
  geom_cladelabel(n13, "Asian fermentation",angle=40, fontsize=3, offset=0.2, hjust=1) + 
  geom_cladelabel(n14, "Taiwanese",angle=0, fontsize=3, offset=0.1, hjust=1) + 
  geom_hilight(node=n15, fill="#2F30EE", alpha=0.45) +
  geom_hilight(node=n16, fill="#FFEE08", alpha=0.45) +
  geom_hilight(node=n14, fill="#999999", alpha=0.5) +
  geom_strip("CIH","ANA", label = "Mosaic region 3", fontsize=3,barsize=0.5, hjust=1) +
  geom_strip("SACE_GAV","CNC",barsize=0.5) 

p2
```



